Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2001-01-01 paper.tex
Remove subsubsection titles in Introduction

% packages
\usepackage{amsmath}              % for \text, align, split ...
\usepackage{amssymb}              % for \gtrsim ...
\usepackage{braket}               % for \bra \ket
\usepackage[mathscr]{euscript}    % for \mathscr
\usepackage{graphicx}             % for \includegraphics
\usepackage{hyperref}             % for \url and \href
\usepackage{listings}             % for typesetting code with \lstinline
\usepackage{tabularx}             % for tabularx

@@

@@

\author{Y.\ L.\ Zhu}
\affiliation{Department of Physics,
  North Carolina State University, Raleigh, NC 27695, USA}

@@

\begin{abstract}
  We present a covariant ray tracing algorithm for computing high-resolution
  neutrino distributions in general relativistic numerical spacetimes with
  hydrodynamical sources.
  Our formulation treats the very important effect of
  elastic scattering of neutrinos off of nuclei and nucleons
  (changing the neutrino's direction but not energy)
  by incorporating estimates of the background neutrino fields.
  These background fields may be taken from a low-order moment simulation
  or be ignored, in which case the method
  reduces to a standard state-of-the-art ray tracing formulation.
  It handles radiation in regimes spanning optically thick to optically thin.
  We test the new code, highlight its strengths and weaknesses,
  and apply it to a simulation of a neutron star merger
  to compute neutrino fluxes and spectra,
  and to demonstrate a neutrino flavor oscillation calculation.
\end{abstract}

@@

\section{Introduction}
Neutrinos are one of the dominant energy transport phenomena at play in
neutron star mergers: heating, cooling, and pushing the disrupted nuclear
matter.
In addition, they change the composition of the matter via charged current
interactions.
Because neutrinos scatter over length scales both large and small with
respect to fluid scales, accurate models require a neutrino treatment that
respects the freedom of neutrino distribution functions to vary
drastically from geometrically simple distributions in thermodynamic equilibrium.

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries, so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{ruff1996-leakage_part1,ross2003-leakage_part2,
  deat2013-leakage, pere2016-asl, radi2016-dynamical}
\footnote{TODO: cite others}
capture some of the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
\footnote{TODO: discuss mc more clearly}
Monte Carlo methods like those
used in supernova simulations \cite{abdi2012-monte_carlo}
and stationary models of accretion disks \cite{rich2015-monte_carlo}
are an excellent tool but require large computational resources
because of their global treatment of the problem.

The state of the art today for neutron star merger calculations
couples radiation and matter using a truncated moment formalism
\citep{fouc2015-m1_nsbh, fouc2016-m1_nsns},
evolving the zeroth- and first-angular moments of the energy density,
that is the total energy- and momentum-densities.
This method, commonly called M1 transport, was formulated by
\cite{thor1981-truncated_moment} and modernized by
\cite{shib2011-truncated_moment}.
M1 transport has recently become popular in the core-collapse supernova
community as well
\cite{kuro2012-m1_ccsne,just2015-m1_code,ocon2015-gr1d_with_nu,
  skin2016-ray_vs_m1,robe2016-gr_m1_ccsne}.

For merger simulations, \cite{fouc2016-m1_evolve_n} have recently
expanded their M1 transport code
to also evolve the zeroth-angular moment of the number density,
providing total energy densities and average energies throughout the
simulated volume.
Even so, with M1 transport codes only evolving two angular moments
and one or two energy moments,
they can extract only limited angular and spectral information
from the neutrino fields.

%\subsubsection*{What physics problems do we want to address?}
But many interesting unsolved problems require an accurate model of the neutrino
spectra and angular distributions.
With a model of the neutrino emission from a merger we can
1) examine neutrino effects on the nucleosynthesis of the ejected material
\citep{surm2011-nickel_56, robe2016-sph_nu_nucleo},
2) explore the rich flavor oscillation physics likely to occur
\citep{malk2012-mnr_1, malk2015-mnr_2, malk2016-mnr_3, zhu2016-mnr_nsns_remnant,
  vaan2016-uncovering_mnr},
3) improve closure relations used in M1 transport schemes
\citep{ramp2002-truncated_moment, shib2011-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}, and
4) study possible jet formation due to neutrino annihilation
\citep{ruff1999-nunubar_nsns, asan2000-nunubar, birk2007-nunubar,
  hari2010-gr_nunubar_collapsar, zala2011-nunubar, leng2014-nunubar,
  just2016-nsns_ejecta_obstacles}.

%\subsubsection*{Why ray tracing?}
Angular and spectral neutrino distributions in neutron star merger simulations may
be available soon, for example with the coupled Monte-Carlo-M1 scheme presented in
\citep{fouc2017-m1_with_mc}.
In this work, however, we present a ray tracing method to compute neutrino
distribution functions from the existing state-of-the-art general relativistic M1
transport hydrodynamics simulations.
We choose ray tracing because it is conceptually simple,
numerically inexpensive,
and extends to high resolution in energy and angle
by simply increasing the number of rays sampled.
Furthermore, the computational implementation parallellizes trivially.

With a ray tracing method we approach radiation transport from the perspective
of a single observer at a spacetime event $x_o^\alpha$.
Our goal is to compute the distribution function
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$, or the amount of neutrino radiation
of species
$\nu_\sigma\in\{\nu_e,\nu_\mu,\nu_\tau,\bar{\nu}_e,\bar{\nu}_\mu,\bar{\nu}_\tau\}$
with momentum $p_\beta$ impinging on $x_o^\alpha$.
To do so we trace a geodesic trajectory from $x_o^\alpha$ in the backwards
direction $-p_\beta$ to sample the incoming radiation along that line of sight.
By tracing a family of rays intersecting $x_o^\alpha$ we build up a
picture of the distribution function there.
And by sampling many observation points we construct a global picture
of $f^{\nu_\sigma}(x^\alpha;p_\beta)$.

The ray tracing framework
is conceptually simple because it solves the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann}) along characteristics,
reducing it to a one-dimensional ordinary differential equation.
It is numerically cheap because it confines computations of
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
with the history of that light-cone truncated at large optical depth.
\footnote{TODO: state scaling $\mathscr{O}(n)$?}
It easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.
And it parallelizes trivially by ray, since each ray is computed
independently.

%\subsubsection*{What is new and better about this ray tracing formulation?}
Several ray tracing formulations for radiation transport already exist.
Most formulations assume an analytic spacetime metric
\citep{birk2007-nunubar, caba2009-detecting_grb_nu,
  hari2010-gr_nunubar_collapsar, kova2011-gr_ray_tracing}.
And many make the simplifying assumption of blackbody emission from a
neutrinosurface
\citep{birk2007-nunubar, caba2009-detecting_grb_nu, kova2011-gr_ray_tracing},
limiting them to equilibrium, optically-thick configurations.
Current state-of-the-art ray tracing formulations avoid the assumption of
blackbody spectra by integrating a local emissivity along each geodesic
(e.g. \cite{hari2010-gr_nunubar_collapsar} for neutrinos and
\cite{youn2012-gr_radiative_transfer} for photons).
But no formulations to date account for the important scattering and pair
processes outlined in Tab.~\ref{tab:neutrino_processes}.

@@

%\subsubsection*{Why does a covariant formulation matter?}
We formulate the ray tracing equations covariantly---free from
assumptions about the spacetime geometry or coordinates. This is essential
because we want to apply the method as a postprocessing step using
time snapshots of data computed from general relativistic evolutions.
The spacetime represented in these snapshots is not analytic (i.e. Kerr).
And even in configurations that are described well by the Kerr metric
(e.g. a low-mass disk around a massive black hole),
the evolution coordinates are unlikely to present the metric in
a familiar analytic form.
This is because integrating the Einstein Equations often requires complicated,
time-dependent gauge conditions
\citep{lind2007-gen_harmonic, fouc2013-compactness_and_spin}.

%\subsubsection*{Why do scattering and pair processes matter?}
Elastic scattering (see Tab.~\ref{tab:neutrino_processes}) can signicantly
modify neutrino distributions in angle and dilute the emitted spectrum over a
larger emitting surface \citep{pere2016-asl}.
This is especially pronounced in the case of heavy-lepton neutrinos.
\footnote{TODO: cite}
Inelastic scattering and pair processes can introduce further modifications.
Any phenomena that involve neutrino-neutrino interactions,
for example neutrino oscillations and neutrino-antineutrino annihilation,
\footnote{TODO: cite}
depend sensitively on the angular distribution.
And spectral changes can strongly affect the nucleosynthesis
occuring in the ejected and irradiated material.
\footnote{TODO: justify these statements}
The ray tracing method we present in this paper captures the dominant effects
of elastic scattering, while leaving the physics of inelastic scattering and
pair processes for later work.

%\subsubsection*{In what ways is this method limited?}
Ray tracing is ideally suited to problems requiring detailed knowledge of
radiation distribution functions over small regions of spacetime:
for example along a matter or radiation trajectory,
or over a small volume outside a source.

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
\footnote{TODO: clarify how momenta are decoupled}
When solving for $f^{\nu_\sigma}(x^\alpha;p_\beta)$,
we have no information about the distribution at different momenta
$f^{\nu_\sigma}(x^\alpha;p'_\beta)$,
or about the distribution of the relevant antineutrino species
$f^{\bar{\nu}_\sigma}(x^\alpha;p'_\beta)$;
these missing data are essential ingredients
for the source terms of the Boltzmann Equation describing the creation and
destruction of neutrinos due to the interaction processes described in
Tab.~\ref{tab:neutrino_processes}.

@@

In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present an application
in the dynamical environment following the merger of two neutron stars,
computing neutrino flavor oscillation along an outgoing ray,
including the effects of coherent forward scattering with ambient neutrinos.
In Sec.~\ref{sec:conclusions} we summarize our work and anticipate improvements.

Greek tensor indices ($\alpha, \beta, ...$) range over all four coordinates,
whereas Latin indices ($i, j, ...$) range over the spatial coordinates 1--3.
We use naturalized units in which $\{\hbar,c,k_{\rm B}\}=1$.
And for most of the remainder of this article we suppress neutrino species label
$\nu_\sigma$ since the formulation is general to any species.
Where we do reference particular species we use the three categories
relevant to the energy scales of mergers, $\nu_e$, $\bar{\nu}_e$, and
$\nu_x=\left\{\nu_\mu,\bar{\nu}_\mu,\nu_\tau,\bar{\nu}_\tau\right\}$.

@@

Because $\ell_\beta$ is subject to two constraints
(normalization and orthogonality to the observer's velocity)
it has two degrees of freedom; we make this explicit by defining its
spatial cartesian components with respect to spherical polar angles
\begin{equation}
  \label{eqn:def_direction}
  \ell_\alpha \rightarrow
  q (s,\sin a \cos b,\sin a\sin b,\cos a),
\end{equation}
with $q$ and $s$ functions of $a$ and $b$.
Now our symbol for the distribution function,
$f(x^\alpha;\varepsilon,a,b)$,
makes manifest its seven independent arguments.

We also define a rotated frame
\begin{equation}
  \ell_{\alpha'}=\frac{\partial x^\alpha}{\partial x^{\alpha'}}\,\ell_\alpha \nonumber
\end{equation}
in which the cartesian components of the direction 1-form are defined
\begin{equation}
  \label{eqn:def_direction_primed}
  \ell_{\alpha'} \rightarrow
  q (s,\sin A \cos B,\sin A\sin B,\cos A),
\end{equation}
so that momenta with vanishing polar angle $A$ move outward along coordinate
radial rays.
This transformation is chosen so that for an observer far from the source,
incoming radiation will be concentrated into a narrow beam around $\cos A\approx1$,
independent of that observer's position in coordinate space.
For explicit definitions see App.~\ref{sec:definitions}.

In Minkowski spacetime, and for a stationary observer
$u_\beta\rightarrow(-1,0,0,0)$, we would have $s=0$, $q=1$,
and $\varepsilon=-p_t$. In that case $\cos A$ may be identified
\footnote{TODO: cite someone that uses $\mu$}
with the familiar forward direction cosine $\mu$.

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\footnote{TODO: point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x} below)
and $C[f]$ is the source term arising from interactions with the medium.
By Eqn.~\ref{eqn:geodesic_x},
the affine parameter has dimension ${\rm length}\,{\rm energy}^{-1}$,
so the source term has dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

We make the right hand side of Eqn.~\ref{eqn:boltzmann} explicit by writing
the source term linear in $f$:
\begin{align}
  \label{eqn:boltzmann_linear}
  \frac{d}{d\lambda}f &=
  \mathscr{E} - \mathscr{K} f \\
  \label{eqn:boltzmann_linear_s}
  &= \mathscr{K}(\mathscr{S}-f)
\end{align}
where we have introduced
$\mathscr{E}$, the invariant total emissivity, and
$\mathscr{K}$, the invariant total opacity.
These describe respectively the energy gained and the energy lost
per length traveled by the neutrino, true scalar quantities
which take identical values for all observers.
In the second form we have introduced the source function
$\mathscr{S}\equiv\mathscr{E}/\mathscr{K}$, which makes manifest the
behavior of the right hand side, driving $f$ toward $\mathscr{S}$
over a lengthscale $\mathscr{K}^{-1}$ in the affine parameter;
or from Eqn.~\ref{eqn:proper_length} below, over a proper lengthscale
$\varepsilon/\mathscr{K}$ measured by the fiducial observer.

These coefficients are computed by considering their dependence
on neutrino and antineutrino distribution functions at other momenta
(i.e.\ Fermi-blocking).
We consider the two dominant classes of interactions in this work:
the absorption/emission (AE) and elastic scattering (SE) processes
listed in Tab.~\ref{tab:neutrino_processes}.
Thus we separate these coefficients:
\begin{align}
  \label{eqn:inv_emissivity_components}
  \mathscr{E} &= \mathscr{E}_{\rm AE} + \mathscr{E}_{\rm SE},\\
  \label{eqn:inv_opacity_components}
  \mathscr{K} &= \mathscr{K}_{\rm AE} + \mathscr{K}_{\rm SE}.
\end{align}

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
for the reactions
\begin{align}
  \nu_e + n            &\leftrightarrow e^- + p,\nonumber\\
  \bar{\nu}_e + p      &\leftrightarrow e^+ + n,\nonumber\\
  \nu_e + {}^AZ        &\leftrightarrow e^- + {}^A(Z+1),\nonumber\\
  \bar{\nu}_e + {}^AZ  &\leftrightarrow e^+ + {}^A(Z-1),\nonumber
\end{align}
with ${}^AZ$ representing a nucleus of mass number $A$ and charge $Z$.
In terms of the emissivity $j(\varepsilon)$ describing number of neutrinos
of energy $\varepsilon$ emitted per length,
and the absorption opacity $\chi_a(\varepsilon)$
describing the number absorbed per length,
the coefficients are
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the distribution function of neutrinos in radiative
equilibrium with the matter, i.e. the Fermi-Dirac distribution function
\begin{equation}
  \label{eqn:feq}
  f^{\rm eq}(\varepsilon) \equiv
  \left(1+e^{\varepsilon/(k_{\rm B} T) -\eta_\nu}\right)^{-1},
\end{equation}
with the neutrino chemical potentials dependent on the local density,
temperature, and composition of the fluid via the neutron, proton,
and electron chemical potentials:
$\eta_{\nu_e}=-\eta_{\bar{\nu}_e}=\eta_p-\eta_n+\eta_{e^-}$ and
$\eta_{\nu_x}=0$.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption \cite{burr2006-neutrino_opacities}.
By detailed balance of the absorption/emission reactions, we may alternatively
write the emissivity in terms of the equilibrium distribution function:
\begin{equation}
  \label{eqn:ae_emissivity_feq}
  \mathscr{E}_{\rm AE}(\varepsilon)
  = \mathscr{K}_{\rm AE}(\varepsilon) f^{\rm eq}(\varepsilon).
\end{equation}
See App.~\ref{ssec:sources_ae} for details.
\footnote{TODO: identify $\mathscr{K}_{\rm AE}$ and $\kappa^*$,
  in e.g.\ \cite{hari2010-gr_nunubar_collapsar}}

The elastic scattering coefficients are computed from
a background field, and a sum over opacities for
the reactions
\begin{align}
  \nu + N     &\leftrightarrow \nu + N,\nonumber\\
  \nu + {}^AZ &\leftrightarrow \nu + {}^AZ,\nonumber
\end{align}
with $N$ standing in for either $n$ or $p$.
In terms of a background field $\Phi(\varepsilon)$ describing
the number of neutrinos of energy $\varepsilon$ present at this event,
and the scattering opacity $\chi(\varepsilon)$ describing the
number scattered to other directions per length,
the coefficients are
\begin{align}
  \label{eqn:se_emissivity}
  \mathscr{E}_{\rm SE}(\varepsilon)
  &= \mathscr{K}_{\rm SE}(\varepsilon) \, \Phi(\varepsilon) \, \\
  \label{eqn:se_opacity_summed}
  \mathscr{K}_{\rm SE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm s},i}(\varepsilon).
\end{align}
In the isotropic limit of trapped radiation,
$\Phi$ is equivalent to $f^{\rm eq}$;
in the free-streaming limit at a distance $r$ from a source,
$\Phi$ attenuates as $r^{-2}$.
See App.~\ref{ssec:sources_se} for details.

>>>>

% packages
\usepackage{amssymb}              % for \gtrsim ...
\usepackage{amsmath}              % for \text, align, split ...
\usepackage{color}                % for colorful margin notes
\usepackage[mathscr]{euscript}    % for \mathscr
\usepackage{float}                % for better control of figure placement
\usepackage{graphicx}             % for \includegraphics
\usepackage{hyperref}             % for \url and \href
\usepackage{listings}             % for typesetting code with \lstinline
\usepackage{tabularx}             % for tabularx

@@

\newcommand{\todo}[1]{\marginpar{\tiny{\textcolor{red}{#1}}}}
%\renewcommand\todo[1]{} % if you wish to hide all todo notes

% force all figures to be placed right where they are in the text
\floatplacement{figure}{H}

@@

@@

\begin{abstract}
  We present a covariant ray tracing algorithm for computing high-resolution
  neutrino distributions in general numerical spacetimes with hydrodynamical
  sources.
  Our formulation treats the very important effect of
  elastic scattering of neutrinos off of nuclei and nucleons
  (changing the neutrino's direction but not energy)
  by incorporating estimates of the background neutrino fields.
  These background fields may be taken from a low-order moment simulation
  or be ignored entirely, in which case the method
  reduces to a standard state-of-the-art ray tracing formulation.
  It handles radiation in regimes spanning optically thick to optically thin.
  We test the new code, highlight its strengths and weaknesses, and
  apply it to simulations of neutron star mergers to
  compute neutrino fluxes as a function of energy and emission angle.
\end{abstract}

@@

\section{Introduction}
Neutrinos are one of the dominant energy transport phenomena at play in
neutron star mergers, heating, cooling, and pushing the disrupted nuclear
matter.
In addition, they change the composition of the matter via charged current
interactions.
Because neutrinos scatter over length scales both large and small with
respect to fluid scales, accurate models require a neutrino treatment that
respects the freedom of neutrino distribution functions to vary
drastically from simple geometric distributions in thermodynamic equilibrium.

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{deat2013-leakage, pere2016-asl, radi2016-dynamical}
\todo{cite others}
capture the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
\todo{discuss mc more clearly}
Monte Carlo methods \citep{abdi2012-monte_carlo, rich2015-monte_carlo}
are an excellent tool require large computational resources because of
their global treatment of the problem.

The state of the art today couples radiation and matter with a truncated
moment formalism
\citep{shib2011-truncated_moment, fouc2015-m1_nsbh, fouc2016-m1_nsns,
  just2015-m1_code},
\todo{cite others}
evolving the zeroth- and first-angular moments of the distribution function
(the energy density and flux respectively) representing the zeroth-energy moment
of the radiation field (the total energy).
This method, commonly called M1 transport, was formulated by
\cite{thor1981-truncated_moment, shib2011-truncated_moment}
and has been applied for example in \cite{ramp2002-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}.
Recently, \cite{fouc2016-m1_evolve_n} have expanded their M1 transport code
to also evolve the number densities of neutrinos,
providing total and average energies throughout the simulated volume.
Even so, with M1 transport codes only evolving two angular moments
and one or two energy moments,
we can extract only limited angular and spectral information
from the neutrino fields.

%\subsubsection*{What physics problems do we want to address?}
But many interesting unsolved problems require an accurate model of the neutrino
spectra and angular distributions.
With a model of the neutrino emission from a merger we can
1) examine neutrino effects on the nucleosynthesis of the ejected material
\citep{surm2011-nickel_56, robe2016-sph_nu_nucleo},
2) explore the rich flavor oscillation physics likely to occur
\citep{malk2012-mnr_1, malk2015-mnr_2, malk2016-mnr_3, zhu2016-mnr_nsns_remnant,
  vaan2016-uncovering_mnr},
3) improve closure relations used in M1 transport schemes
\citep{ramp2002-truncated_moment, shib2011-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}, and
4) study possible jet formation due to neutrino annihilation
\citep{ruff1999-nunubar_nsns, asan2000-nunubar, birk2007-nunubar,
  hari2010-gr_nunubar_collapsar, zala2011-nunubar, leng2014-nunubar}.

%\subsubsection*{Why ray tracing?}
In this work we present a ray tracing method to compute neutrino distribution
functions from state-of-the-art general relativistic M1 transport
hydrodynamics simulations.
We choose ray tracing because it is conceptually simple, numerically cheap,
and easily extends to high resolution in energy and angle.

With a ray tracing method we approach radiation transport from the perspective
of a single observer at a spacetime event $x_o^\alpha$.
Our goal is to compute the distribution function
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$, or the amount of neutrino radiation
of species $\nu_\sigma\in\{\nu_{\{e,\mu,\tau\}},\bar{\nu}_{\{e,\mu,\tau\}}\}$
with momentum $p_\beta$ impinging on $x_o^\alpha$.
To do so we trace a geodesic trajectory from $x_o^\alpha$ in the backwards
direction $-p_\beta$ to sample the incoming radiation along that line of sight.
By tracing a family of rays intersecting $x_o^\alpha$ we build up a
picture of the distribution function there.
And by sampling many observation points we construct a global picture
of $f^{\nu_\sigma}(x^\alpha;p_\beta)$.

The ray tracing framework
is conceptually simple because it solves the equation of radiation
transport (Eqn.~\ref{eqn:boltzmann}) along characteristics,
reducing it to a one-dimensional ordinary differential equation.
It is numerically cheap because it confines computations of
$f^{\nu_\sigma}(x_o^\alpha;p_\beta)$ to the past light-cone of $x_o^\alpha$,
with the history of that light-cone truncated at large optical depth.
\todo{state scaling $\mathscr{O}(n)$?}
And it easily extends to high resolution in energy and angle by simply
increasing the number of rays sampled.

%\subsubsection*{What is new and better about this ray tracing formulation?}
Several ray tracing formulations for radiation transport already exist.
Most formulations assume an analytic spacetime metric
\citep{birk2007-nunubar, hari2010-gr_nunubar_collapsar, kova2011-gr_ray_tracing}.
And many make the simplifying assumption of blackbody emission from a
neutrinosurface \citep{birk2007-nunubar, kova2011-gr_ray_tracing},
limiting them to equilibrium, optically thick configurations.
Current state-of-the-art ray tracing formulations avoid the assumption of
blackbody spectra by integrating a local emissivity along each geodesic
(e.g. \cite{hari2010-gr_nunubar_collapsar} for neutrinos and
\cite{youn2012-gr_radiative_transfer} for photons).
But no formulations to date account for the important scattering and pair
processes outlined in Tab.~\ref{tab:neutrino_processes}.

@@

%\subsubsection*{Why does a covariant formulation matter?}
We formulate our ray tracing equations covariantly---free from
assumptions about the spacetime geometry or coordinates. This is essential
because we want to apply the method as a postprocessing step using
time snapshots of data computed from general relativistic evolutions.
The spacetime represented in these snapshots is not analytic (i.e. Kerr).
And even in configurations that are described well by the Kerr metric
(e.g. a low-mass disk around a massive black hole),
the evolution coordinates are unlikely to present the metric in
a familiar analytic form.
This is because integrating the Einstein Equations often requires complicated,
time-dependent gauge conditions
\citep{lind2007-gen_harmonic, fouc2013-compactness_and_spin}.

%\subsubsection*{Why do scattering and pair processes matter?}
Elastic scattering (see Tab.~\ref{tab:neutrino_processes}) can signicantly
modify neutrino distributions in angle and dilute the emitted spectrum over a
larger emitting surface \citep{pere2016-asl}.
This is especially pronounced in the case of heavy-lepton neutrinos.
\todo{cite}
Inelastic scattering and pair processes can introduce further modifications.
Any phenomena that involve neutrino-neutrino interactions,
for example neutrino oscillations and neutrino-antineutrino annihilation,
\todo{cite}
depend sensitively on the angular distribution.
And spectral changes can strongly affect the nucleosynthesis
occuring in the ejected and irradiated material.
\todo{justify these statements}
The ray tracing method we present in this paper captures the dominant effects
of elastic scattering, while leaving the physics of inelastic scattering and
pair processes for later work.

%\subsubsection*{In what ways is this method limited?}
Ray tracing is ideally suited to problems requiring detailed knowledge of
radiation distribution functions over small regions of spacetime:
for example along a matter or radiation trajectory,
or over a small volume outside a source.
But the method becomes computationally burdensome when the inquiry extends
to large volumes of spacetime.

More fundamentally, though, ray tracing is limited by its naive
treatment of the Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}),
a treatment which essentially decouples different neutrino momenta and species.
\todo{clarify how momenta are decoupled}
When we're solving for $f^{\nu_\sigma}(x^\alpha;p_\beta)$, we have no
information about $f^{\nu_\sigma}(x^\alpha;p'_\beta)$, and
$f^{\bar{\nu}_\sigma}(x^\alpha;p'_\beta)$, which are essential ingredients
for the source terms of the Boltzmann Equation describing the creation and
destruction of neutrinos due to the interaction processes described in
Tab.~\ref{tab:neutrino_processes}.

@@

\subsubsection*{Outline}
In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present several applications
in the dynamical environment following the merger of two neutron stars:
computing neutrino spectra as a function of observer's angle,
computing the self-interaction potential descriptive of flavor evolution
in high-neutrino-density environments,
and computing the energy dumped into jet-forming electron-positron
pairs via neutrino-antineutrino annihilation.

Greek tensor indices ($\alpha, \beta, ...$) range over all four coordinates,
whereas Latin indices ($i, j, ...$) range over the spatial coordinates 1--3.
We use naturalized units in which $\hbar=c=1$.
And for the remainder of this article we suppress the neutrino species label
$\nu_\sigma$ since the formulation is general to any species.

@@

Because $\ell_\beta$ is subject to two constraints
(normalization and orthogonality to the observer's velocity)
it has two degrees of freedom; we make this explicit by defining its
spatial components with respect to spherical polar angles
\begin{equation}
  \label{eqn:def_direction}
  \ell_\alpha :=
  q (s,\sin\upsilon_1\cos\upsilon_2,\sin\upsilon_1\sin\upsilon_2,\cos\upsilon_1),
\end{equation}
with $q$ and $s$ functions of $\upsilon_1$ and $\upsilon_2$.
Now our symbol for the distribution function,
$f(x^\alpha;\varepsilon,\upsilon_1,\upsilon_2)$,
makes manifest its seven independent arguments.
\todo{also decompose wrt $p_t$}

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium,
having dimension ${\rm energy}\,{\rm length}^{-1}$.
\todo{describe $C[f]$ in physical terms}
The source term varies over phase space $(x^\alpha,p_\beta)$,
and due to Fermi blocking depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

We make the right hand side of Eqn.~\ref{eqn:boltzmann} explicit by writing
the source term linear in $f$:
\begin{equation}
  \label{eqn:boltzmann_linear}
  \frac{d}{d\lambda}f =
  \mathscr{E} - \mathscr{K} f,
\end{equation}
where we have introduced
\todo{describe $\mathscr{E}$ and $\mathscr{K}$ in physical terms}
$\mathscr{E}$, the invariant total emissivity, and
$\mathscr{K}$, the invariant total opacity.
These coefficients are computed by considering their dependence
on neutrino and antineutrino distribution functions at other momenta
(i.e.\ Fermi-blocking).

We consider two classes of interactions in this work:
the absorption/emission (AE) and elastic scattering (SE) processes
listed in Tab.~\ref{tab:neutrino_processes}.
Thus we separate these coefficients:
\begin{align}
  \label{eqn:inv_emissivity_components}
  \mathscr{E} &= \mathscr{E}_{\rm AE} + \mathscr{E}_{\rm SE},\\
  \label{eqn:inv_opacity_components}
  \mathscr{K} &= \mathscr{K}_{\rm AE} + \mathscr{K}_{\rm SE}.
\end{align}

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
in Tab.~\ref{tab:neutrino_processes}:
\todo{define $j$ and $\chi$}
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the Fermi-Dirac distribution function with the chemical
potential dependent on the local density, temperature, and composition of the
fluid.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption.
See App.~\ref{ssec:sources_ae} for details.
\todo{identify $\mathscr{K}_{\rm AE}$ and $\kappa^*$,
  in e.g.\ \cite{hari2010-gr_nunubar_collapsar}}

The elastic scattering coefficients are computed from
a background field, and a sum over opacities for
the relevant reactions in Tab.~\ref{tab:neutrino_processes}:
\begin{align}
  \label{eqn:se_emissivity_summed}
  \mathscr{E}_{\rm SE}(\varepsilon)
  &= \mathscr{K}_{\rm SE}(\varepsilon) \, \Phi_{\rm SE}(\varepsilon) \, \\
  \label{eqn:se_opacity_summed}
  \mathscr{K}_{\rm SE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm s},i}(\varepsilon).
\end{align}
See App.~\ref{ssec:sources_se} for details.

This paper introduces the use of neutrino densities and fluxes evolved in an
M1 transport simulation to estimate the background field, $\Phi_{\rm SE}$.
App.~\ref{ssec:sources_se} details a method to calculate
$\Phi_{\rm SE}(\varepsilon)$ in two different ways:
\begin{itemize}
\item
  the \emph{spectral} method using densities and fluxes
  extracted from a simulation evolved over multiple energy groups
  to compute the background field with Eqn.~\ref{eqn:se_factor_expansion},
\item
  the \emph{gray} method using energy-integrated densities and fluxes
  extracted from a gray simulation and approximating the
  energy distribution with Eqns.~\ref{eqn:J_from_gray} and
  \ref{eqn:H_from_gray}.
\end{itemize}
